### 2020-06-03 report/sections/server/node/efsm.tex
Report: complete EFSM ssubsection for chapter 4

\subparagraph{Send}
We model selections using a union type to
encapsulate the possible send actions, as shown in 
\cref{lst:addersvrsendhandler}.
Each send action is encoded as a tuple of
the label, the payload, and the successor state encoding.
We see some benefits from defining Message Types as interfaces:
with deterministic properties, we can index into the message type
to access the property's type, 
e.g. \lstonelinejs{Message.S54THANKS['payload']} 
would resolve to
\lstonelinejs{[string]},
based on \cref{lst:addersvrmsg}.

@@

@@

\begin{itemize}
\item need a way for the runtime to distinguish between states (implementation)
\item attempt at using conditional types, but because they handle conditional union types in a distributed way (give example), cannot be statically typed
\item alternative -- use discriminated union by wrapping in an Implementation class
\item all non-terminal states will `return' the successor implementation -- very difficult to resolve in the type-checker in the runtime (give example of what it may look like) -- solve by giving each state implementation the `advance' function to respect the event-driven nature of everything, so they can call it on completion
\item visualise the interaction between runtime and states as a sequence diagram for message passing
\end{itemize}

>>>>

\subparagraph{Send}
We model selections using a union type to
encapsulate the possible send actions, as shown in 
\cref{lst:addersvrsendhandler}.
Each send action is encoded as a tuple of
the label, the payload, and the successor state encoding.
We see some benefits from defining Message Types as interfaces:
TypeScript supports \textit{index type queries} to extract
named property types, so
\lstonelinejs{Message.S54THANKS['payload']} 
would resolve to \lstonelinejs{[string]},
based on the interface definition \cref{lst:addersvrmsg}.

@@

\subsubsection{Distinguishing Handlers using Conditional Types}

@@

We try to define type dependencies using
\textit{conditional types} in TypeScript.
A conditional types is a type-level expression
\begin{lstlisting}[language=javascript,numbers=none]
T extends U ? X : Y;
\end{lstlisting}
which reads, \textit{if \texttt{T} is assignable to \texttt{U},
then the type is \texttt{X}; otherwise, the type is \texttt{Y}}.

Combined with \textit{generic constraints}\footnote{
\lstonelinejs{<T extends U>} defines a generic type \texttt{T}
and enforces that it must be a type assignable to \texttt{U}.
},
we can approximate the dependency between the state identifier
enum and the generated handler API using something
similar to \cref{lst:conditionaltypes}.

\begin{figure}[!h]
\begin{lstlisting}[language=javascript,tabsize=2]
enum SendState { S1, S3, ... };
enum ReceiveState { S2, S4, ... };
type State = SendState | ReceiveState;

type SendHandler<S extends SendState> = 
	S extends SendState.S1 ? Handler.S1 :
	S extends SendState.S3 ? Handler.S3 : ... ;

type ReceiveHandler<S extends ReceiveState> = 
	S extends ReceiveState.S2 ? Handler.S2 :
	S extends ReceiveState.S4 ? Handler.S4 : ... ;
\end{lstlisting}
\captionof{lstlisting}{Approximating Type Dependency
using \textit{Conditional Types}}
\label{lst:conditionaltypes}
\end{figure}

We intend to use this construct when defining the EFSM transition
function for the runtime, for each type of state,
so the method signature for transitioning to send states
would resemble \cref{lst:conditionaltransitionfunction}

\begin{figure}[!h]
\begin{lstlisting}[language=javascript,numbers=none]
declare function transitionToSend<S extends SendState>(
	stateId: S, handler: SendHandler<S>
);
\end{lstlisting}
\captionof{lstlisting}{EFSM Transition Function 
using Conditional Types}
\label{lst:conditionaltransitionfunction}
\end{figure}

Unfortunately, this approach does not work for the simple fact that
conditional types were not designed to be exploited in this manner.
The main limitation of conditional types is its 
\textit{distributivity} when the type ``parameter'' is an
union type (which is the case for enums, as
\texttt{S = SendState.S1 | SendState.S3 | ...}), where

\begin{lstlisting}[language=javascript,numbers=none]
(T1 | T2) extends U ? X : Y
\end{lstlisting}

results in the conditional type being \textit{distributed}
among each constituent,

\begin{lstlisting}[language=javascript,numbers=none]
(T1 extends U ? X : Y) | (T2 extends U ? X : Y)
\end{lstlisting}

so the type expression returns to an union type,

\begin{lstlisting}[language=javascript,numbers=none]
X | Y
\end{lstlisting}

Returning to \cref{lst:conditionaltransitionfunction}, 
the type of \texttt{handler} will end up being a union type, 
rather than the ``dependent type''
construct we were hoping for.

\subsubsection{Distinguishing Handlers using Discriminated Unions}

Instead, we leverage \textit{discriminated unions}: all members
of the union type share a common property (the \textit{discriminant})
of which they each define an
unique value for, so that the TypeScript Compiler can refine the union
to the specific constituent upon checking the value of the discriminant
(e.g. applying a switch statement).

For the time being,
it is sufficient to understand that for each EFSM state,
in addition to the API defined under the
\texttt{Handler} namespace, it also has a wrapper API defined under
the \texttt{Implementation} namespace (\cref{lst:nodeefsmimplementation}),
which defines the \texttt{type}
discriminant property internally. This explains why the successor
state encodings in 
\cref{lst:addersvrsendhandler,lst:addersvrreceivehandler}
were defined as such.

\begin{figure}[!h]
\begin{lstlisting}[language=javascript,tabsize=2]
abstract class ISend { type: 'Send' = 'Send'; ... }
abstract class IReceive { type: 'Receive' = 'Receive'; ... }

export namespace Implementation {

	export class S51 extends IReceive {
		constructor(private handler: Handler.S51) { super(); }
		...
	}
	...
};
\end{lstlisting}
\captionof{lstlisting}{Discriminated Unions in EFSM 
for Server-Side Endpoints}
\label{lst:nodeefsmimplementation}
\end{figure}

We discuss our runtime implementation shortly 
(\cref{section:noderuntime}), where we disclose more details regarding
the role of the \texttt{Implemenation} wrapper API in
the runtime.

### 2020-06-03 report/sections/server/node/runtime.tex
Report: complete EFSM ssubsection for chapter 4

@@

@@

\subsection{Executing the EFSM}
\begin{itemize}
\item implementation details delegated to the actual implementation states
\item constructor binds all methods (that will be passed to other components0 to `this' because of how javascript works with respect to the scoping of `this' (give short example -- or ignore, if this is in the background for typescript)
\item advance -- use discriminated union to figure out what to pass to the state (send -- sendMessage, receive -- register)
\end{itemize}

>>>>

We define the session runtime in Svr.ts,
named after the endpoint.
It exposes (through the module export) a public API 
(\cref{lst:noderuntimepublicapi})
for the developer to pass in the WebSocket server and application logic
(i.e. their handler implementations).

@@

\begin{figure}[!h]
\begin{lstlisting}[language=javascript,tabsize=2,title=Svr.ts]
export class Svr {
	constructor(wss: WebSocket.Server,
							initialState: Implementation.S51) { ... }
	...
}
\end{lstlisting}
\captionof{lstlisting}{Public API of Session Runtime for Server Endpoint}
\label{lst:noderuntimepublicapi}
\end{figure}

@@

\subsection{Executing the EFSM}
\begin{itemize}
\item implementation details delegated to the actual implementation states
\item all non-terminal states will `return' the successor implementation -- very difficult to resolve in the type-checker in the runtime (give example of what it may look like) -- solve by giving each state implementation the `advance' function to respect the event-driven nature of everything, so they can call it on completion
\item visualise the interaction between runtime and states as a sequence diagram for message passing
\item constructor binds all methods (that will be passed to other components0 to `this' because of how javascript works with respect to the scoping of `this' (give short example -- or ignore, if this is in the background for typescript)
\item advance -- use discriminated union to figure out what to pass to the state (send -- sendMessage, receive -- register)
\end{itemize}
